On the number of tiles visited by a\\ line segment on a rectangular grid

### 2021-12-05 paper.tex
Rewriting

Theorems~\ref{theo: funt, form} and \ref{theo: funl, form} not only give the solutions $\funt(\len)$ and $\funl(\tiles)$ to the first two problems stated in \S\ref{part: intro}; they also provide a way to actually position a segment of length $\len$ or slightly greater than $\funl(\tiles)$, respectively, so that it visits $\funt(\len)$ or $\tiles$ tiles. Namely, for $\isoli$, $\jsoli$ computed as in the corresponding theorem, the segment should have its endpoints in the interiors of two tiles separated $\isoli-1$ steps horizontally and $\jsoli-1$ steps vertically, with its exact position and orientation  adjusted to avoid any grid points.

@@

\begin{corollary}
\label{theo: funl, sq, form}
For a unit square grid, and for $\tiles \in \mathbb N$,
\begin{equation}
\funl(\tiles) = \begin{cases}
\displaystyle
0 & \text{for } \tiles =1, 2 \\[1.4mm]
\displaystyle
\frac{\tiles-3}{\sqrt{2}} & \text{for } \tiles \text{ odd, } \tiles \geq 3 \\[4.5mm]
\displaystyle
\sqrt{\frac{(\tiles-3)^2+1} {2}} & \text{for } \tiles \text{ even, } \tiles \geq 4,
\end{cases}
\end{equation}
or equivalently
\begin{equation}
\label{eq: theo: funt, sq, form; funl}
\funl(\tiles) = \begin{cases}
\displaystyle
0 & \text{for } \tiles =1, 2 \\[1.4mm]
\displaystyle
\sqrt{\left\lceil \frac{(\tiles-3)^2} {2} \right\rceil} & \text{for } \tiles \geq 3.
\end{cases}
\end{equation}
\end{corollary}

@@

From Theorem~\ref{theo: funt, sq, form} it stems that odd values of $\funt(\len)$ correspond to $\isoli = \jsoli$, whereas even values are achieved with $\jsoli = \isoli+1$. In addition, noting that $\funt(\len) = \tiles$ is equivalent to $\funl(\tiles) < \len \leq \funl(\tiles+1)$ and using Corollary~\ref{theo: funl, sq, form} the following characterization of $\funt$ is obtained. For $\tiles \geq 3$ with $\tiles$ odd, $\funt(\len) = \tiles$ if and only if
\begin{equation}
\label{eq: square, odd max tiles, lengths}
\len \in \left( \frac{\tiles-3}{\sqrt{2}}, \frac{ \sqrt{(\tiles-1)^2+(\tiles-3)^2}}{2} \right].
\end{equation}
For $\tiles \geq 4$, $\tiles$ even, $\funt(\len) = \tiles$ if and only if
\begin{equation}
\label{eq: square, even max tiles, lengths}
\len \in \left(\frac{ \sqrt{(\tiles-2)^2+(\tiles-4)^2}}{2}, \frac{\tiles-2}{\sqrt{2}} \right].
\end{equation}
% *** The corollary says \sqrt{\frac{(\tiles-3)^2+1} {2}}, which comes from the corresponding theorem. This is equal to \frac{\sqrt{(\tiles-2)^2+(\tiles-4)^2}} {2}. Here we are using the latter expression, which is probably needed for the random analysis. If so, the expression in the corollary should be changed from the first to the second. (But this change cannnot be made in the theorem it comes from. It's hopefully not a problem that the expression changes from the theorem to the corollary).

@@

Consider again a grid with horizontal spacing $a$ and vertical spacing $b$. The grid crossings decompose into crossings of horizontal and vertical grid lines. By linearity of expectation, the expected number of crossings is the sum of the expectations for parallel lines with spacing $a$ and $b$ respectively, which gives
\[
\frac{2\len}{\pi}\left(\frac 1 a + \frac 1 b\right).
\]
As noted in Proposition \ref{prop: i+j-1}, the number of tiles visited by a segment is the count of its grid line crossings plus $1$, unless it exactly passes through any grid points, but that occurs with probability $0$ and therefore does not affect the expected value. Thus $\funta(\len)$ is obtained by adding $1$ to the above expression, which gives \eqref{eq: funta, form}.
\end{proof}

@@

\begin{proof}
The probability $\probmax(\len)$ becomes $0$ whenever the length causes the maximum number of visited tiles to increase by $1$. This establishes \eqref{eq: probmax liminf}.

For \eqref{eq: probmax lim odd} it suffices to note that
\begin{align}
\label{eq: tedious limit, odd 1}
\lim_{\tiles \rightarrow \infty}\, \tiles g\left(\frac{\sqrt{(\tiles-3)^2+(\tiles-1)^2}} 2, \tiles-3, \tiles-3 \right) &= \frac 2 {3\pi} \\
\label{eq: tedious limit, odd 2}
\lim_{\tiles \rightarrow \infty}\, \tiles g\left(\frac{\sqrt{(\tiles-3)^2+(\tiles-1)^2}} 2, \tiles-5, \tiles-1 \right)
& = \frac 2 {3\pi},
\end{align}
as can be (somewhat tediously) checked using L'H\^opital's rule \cite[section~5.3]{Abbott15}.
% It's very tedious by hand. I used Maple to compute the limit, and then checked it numerically. Maybe one or two intermediate steps from applying L'H\^opital's rule should be included. Perhaps those can be obtained with Maple too, and this part can be rewritten.
Combining these expressions with the second equality in \eqref{eq: probmax} yields \eqref{eq: probmax lim odd}.

Similarly,
\begin{align}
\label{eq: tedious limit, even 1}
\lim_{\tiles \rightarrow \infty}\, \tiles g\left(\frac{\tiles-2}{\sqrt 2}, \tiles-4, \tiles-2 \right) &= \frac 2 {3\pi} \\
\label{eq: tedious limit, even 2}
\lim_{\tiles \rightarrow \infty}\, \tiles g\left(\frac{\tiles-2}{\sqrt 2}, \tiles-6, \tiles \right)
& = \frac 2 {3\pi},
\end{align}
and combining with the fourth equality in \eqref{eq: probmax} gives \eqref{eq: probmax lim even}.

>>>>

Theorems \ref{theo: funt, form} and \ref{theo: funl, form} not only give the solutions $\funt(\len)$ and $\funl(\tiles)$ to the first two problems stated in \S\ref{part: intro}; they also provide a way to actually position a segment of length $\len$ or slightly greater than $\funl(\tiles)$, respectively, so that it visits $\funt(\len)$ or $\tiles$ tiles. Namely, for $\isoli$, $\jsoli$ computed as in the corresponding theorem, the segment should have its endpoints in the interiors of two tiles separated $\isoli-1$ steps horizontally and $\jsoli-1$ steps vertically, with its exact position and orientation  adjusted to avoid any grid points.

@@

\begin{corollary}
\label{theo: funl, sq, form}
For a unit square grid, and for $\tiles \in \mathbb N$,
\begin{equation}
\funl(\tiles) = \begin{cases}
\displaystyle
0 & \text{for } \tiles =1, 2 \\[1.4mm]
\displaystyle
\frac{\tiles-3}{\sqrt{2}} & \text{for } \tiles \text{ odd, } \tiles \geq 3 \\[4.5mm]
\displaystyle
\frac{\sqrt{(\tiles-4)^2+(\tiles-2)^2}} {2} & \text{for } \tiles \text{ even, } \tiles \geq 4,
\end{cases}
\end{equation}
or equivalently
\begin{equation}
\label{eq: theo: funt, sq, form; funl}
\funl(\tiles) = \begin{cases}
\displaystyle
0 & \text{for } \tiles =1, 2 \\[1.4mm]
\displaystyle
\sqrt{\left\lceil \frac{(\tiles-3)^2} {2} \right\rceil} & \text{for } \tiles \geq 3.
\end{cases}
\end{equation}
\end{corollary}
% Theorem~\ref{theo: funl, form} for $a=b=1$ gives $\sqrt{\frac{(\tiles-3)^2+1} {2}}$. Instead, this corollary uses the equivalent expression $\frac{\sqrt{(\tiles-4)^2+(\tiles-2)^2}} {2}$, which is better for subsequent analysis.

@@

From Theorem~\ref{theo: funt, sq, form} it stems that odd values of $\funt(\len)$ correspond to $\isoli = \jsoli$, whereas even values are achieved with $\jsoli = \isoli+1$. In addition, noting that $\funt(\len) = \tiles$ is equivalent to $\funl(\tiles) < \len \leq \funl(\tiles+1)$ and using Corollary~\ref{theo: funl, sq, form} the following characterization of $\funt$ is obtained. For $\tiles \geq 3$ with $\tiles$ odd, $\funt(\len) = \tiles$ if and only if
\begin{equation}
\label{eq: square, odd max tiles, lengths}
\len \in \left( \frac{\tiles-3}{\sqrt{2}}, \frac{ \sqrt{(\tiles-3)^2+(\tiles-1)^2}}{2} \right].
\end{equation}
For $\tiles \geq 4$, $\tiles$ even, $\funt(\len) = \tiles$ if and only if
\begin{equation}
\label{eq: square, even max tiles, lengths}
\len \in \left(\frac{ \sqrt{(\tiles-4)^2+(\tiles-2)^2}}{2}, \frac{\tiles-2}{\sqrt{2}} \right].
\end{equation}

@@

Consider again a grid with horizontal spacing $a$ and vertical spacing $b$. The grid crossings decompose into crossings of horizontal and vertical grid lines. By linearity of expectation, the expected number of crossings is the sum of the expectations for parallel lines with spacing $a$ and $b$ respectively, which gives
\[
\frac{2\len}{\pi}\left(\frac 1 a + \frac 1 b\right).
\]
As noted in Proposition~\ref{prop: i+j-1}, the number of tiles visited by a segment is the count of its grid line crossings plus $1$, unless it exactly passes through any grid points, but that occurs with probability $0$ and therefore does not affect the expected value. Thus $\funta(\len)$ is obtained by adding $1$ to the above expression, which gives \eqref{eq: funta, form}.
\end{proof}

@@

\begin{proof}
The probability $\probmax(\len)$ becomes $0$ whenever the length causes the maximum number of visited tiles to increase by $1$. This proves \eqref{eq: probmax liminf}.

For \eqref{eq: probmax lim odd} it suffices to note that
\begin{equation}
\begin{split}
\label{eq: tedious limit, odd}
&\lim_{\tiles \rightarrow \infty}\, \tiles g\left(\frac{\sqrt{(\tiles-3)^2+(\tiles-1)^2}} 2, \tiles-3, \tiles-3 \right) \\
&\quad = \lim_{\tiles \rightarrow \infty}\, \tiles g\left(\frac{\sqrt{(\tiles-3)^2+(\tiles-1)^2}} 2, \tiles-5, \tiles-1 \right) = \frac 2 {3\pi},
\end{split}
\end{equation}
as can be (somewhat tediously) checked using L'H\^opital's rule \cite[section~5.3]{Abbott15}.
% It's very tedious by hand. I used Maple to compute the limit, and then checked it numerically. Maybe one or two intermediate steps from applying L'H\^opital's rule should be included. Perhaps those can be obtained with Maple too, and this part can be rewritten.
Combining these expressions with the second equality in \eqref{eq: probmax} yields \eqref{eq: probmax lim odd}.

Similarly,
\begin{equation}
\label{eq: tedious limit, even}
\lim_{\tiles \rightarrow \infty}\, \tiles g\left(\frac{\tiles-2}{\sqrt 2}, \tiles-4, \tiles-2 \right) =
\lim_{\tiles \rightarrow \infty}\, \tiles g\left(\frac{\tiles-2}{\sqrt 2}, \tiles-6, \tiles \right)
= \frac 2 {3\pi},
\end{equation}
and combining with the fourth equality in \eqref{eq: probmax} gives \eqref{eq: probmax lim even}.
